### 2023-07-16 tex/agda-primer.tex
[ writing ] address all remaining todos

\chapter{Introduction}

This thesis advances the frontier of programming languages work that can be done
in a proof assistant.
I will elaborate on both of these aspects in the following paragraphs.

The main aim of programming language research is to find common patterns in
computer programs and represent these patterns via programming language
features.
As a basic example, consider functions in, say, a low level language like C, in
contrast to what these functions are compiled to in assembly language/machine
code.
In order to support function calls, the C runtime system manages a
\emph{call stack}.
A function call is compiled to a sequence of instructions which store the return
address and the values of the arguments on the stack, move the stack pointer,
and jump to the compiled code of the function.
A return from a function then places the return value into the appropriate
register, moves the stack pointer back, and jumps back to the return address.
Advances in programming language techniques are evaluated by the usefulness of
behaviours captured and goodness of the mathematical properties of the
abstractions produced.

Within programming language research, type theory is a methodology for producing
abstractions.
A type system lets us enforce in our language complex invariants which maintain
abstraction boundaries while also ideally being machine-checkable.
For example, we give functions function type to abstract away their
implementation via call stacks and jumps and so on.

In this thesis, the invariants of interest revolve around restricting the usage
of variables.
I introduce this topic thoroughly in \cref{sec:linear}, but in brief\ldots

The work of this thesis relies upon type theory in two distinct ways.
Firstly, the main objects of study are programming languages with interesting
type systems.
Secondly, type theory provides the basis of the proof assistant Agda I use to
implement the aforementioned programming languages and operations upon them.

As for the other topic of this thesis,
a \emph{proof assistant}, also known as an \emph{interactive theorem prover}, is
a piece of software that allows for the encoding of mathematical definitions,
theorems, constructions, and proofs, and furthermore check that such encoded
proofs are correct and that such encoded constructions are well formed.
To truly by interactive, i.e.\ to actually assist, a proof assistant will
usually have a user interface which can read partial proofs, display
information about what more proof needs to be given, and provide actions that
will help complete the proof.

Proof assistants have seen increasing use in programming language research in
recent years.
The most obvious reason why working in a proof assistant is seen as beneficial
is that it ensures correctness.
If the proof assistant accurately implements a suitable mathematical foundation,
then any theorem proved in the proof assistant is guaranteed to be a true
theorem of that foundation.
These guarantees of correctness are particularly important when working with
combinatorially complex mathematical objects, proofs about which often require
the consideration of a large number of cases.
Programming language syntaxes are often such complex objects, motivating the use
of proof assistants when studying programming languages.

A second reason to use proof assistants is for the assistance they provide when
exploring a mathematical theory.
When we make a new definition, we may want to test how it works in a special
case, or what constructions it allows us to perform.
In a proof assistant, the assistance tools give us immediate feedback as to what
moves are and aren't allowed.
For example, if we define a complex type system, a proof assistant will let us
interactively build typing derivations, making clear any side conditions and
types of subderivations as we go.
Also, as I do later in this thesis, a proof assistant allows us to build a very
general theory, and practically use that theory directly in more specific cases
without losing rigour.

Thirdly, analogously to how a strong static type system can give us more
confidence when refactoring a program, the constant checking of proofs in a
proof assistant gives us the confidence to change definitions and lemmas knowing
that we will be guided towards the parts of our theory that need to be
correspondingly changed.
This can help if we are developing a new programming language with a changing
specification.

Finally, many proof assistants --- including Agda~\citep{Agda}, which I use in
this thesis --- double as programming languages themselves.
This means that we can write progams and prove properties of them using the same
tool.
Also, many theorems proven in such a proof assistant have computational content.
For example, if we prove a normalisation theorem for a programming language,
this will typically yield a (verified) normalisation algorithm for it, which we
can really run on a computer.
As such, a development in a proof assistant can provide a reference
implementation of a programming language, or even --- as with Idris
2~\citep{Brady21}, Lean 4~\citep{deMU21}, and Cedille~\citep{GRS16} --- the
actual implementation of a programming language.

FIXME: the start of this chapter is directly from the ESOP22 paper.

In this paper, we treat the metatheory of a class of substructural type
systems related to linear logic~\cite{girard87linear}.
This class is variously known as
\emph{coeffectful}~\cite{PetricekOM14,Granule18},
\emph{quantitative}~\cite{BrunelGMZ14,Atkey18}, or
\emph{resource-aware}~\cite{GhicaS14},
or is given no particular name~\cite{reed10distance,abadi99core},
and generalises bounded linear logic to track variable usage with semiring
annotations.
In all of these systems, we have some ambient semiring $\Ann$, and in the
judgements of the type system, variables are annotated by elements of $\Ann$
describing \emph{how} that variable can be used.
The additive structure of $\Ann$ gives the ability to count, or otherwise
accumulate, usages of variables in multiple subterms.
The multiplicative structure gives rise to a form of modality, for example
allowing multiple or unlimited reuse, or movement between security levels, in
the type system.

The aspect of such systems we tackle here is their basic metatheory and
mechanisation thereof.

We build upon both the general structural framework of
\citet{AACMM21} and the substructural techniques of \citet{WA21}.
The way \citeauthor{AACMM21} consolidate and codify mechanisation techniques for
propositional natural deduction systems based on intrinsically typed syntax and
de Bruijn indices, we aim to replicate for linear-like systems based on
semiring usage annotations.
By picking a trivial semiring, our work can subsume that of
\citeauthor{AACMM21}, except for the many pieces of machinery we have not yet
ported to this new framework.

Our work complements that of \citet{Granule18} on the Granule programming
language.
Where Granule focuses on writing programs \emph{in} the language and running
them, we focus on metatheoretic reasoning \emph{about} type systems.
%As such, whereas Granule has a convenient syntax, performant type-checker,
%interpreter

Our work is similar in scope to that of \citet{LicataSR17}, though we work in
a natural deduction style rather than a sequent calculus style.
Where \citeauthor{LicataSR17} are much more agnostic in terms of
substructurality --- allowing for non-commutative and bunched logics ---
we are much more agnostic in terms of syntax.
The system of \citeauthor{LicataSR17} is essentially a single calculus,
supporting ``product'' ($\mathrm F$) types and ``function'' ($\mathrm U$)
types, parametrised on a \emph{mode theory} describing its structural rules.
For this system, they derive the strong result of cut elimination.
Meanwhile, we leave syntax design to the user, and consequently can only
guarantee substitution (which we can only get because of our commitment to
natural deduction).

\section{Outline of the thesis}

This thesis proceeds as follows.
The next two chapters, \cref{sec:simple,sec:linearity}, are introductory in
nature, and cover two largely independent strands of prior work.
In \cref{sec:simple}, I introduce existing methods of representing and reasoning
about type systems in proof assistants based on dependent type theory.
I start from well established representations of well scoped and well typed
terms, and develop these towards a recent approach to environment-based
semantics given by \citet{AACMM21}.
In \cref{sec:linearity}, I discuss the challenges faced when one extends a
treatment of a simple type system, such as that given in \cref{sec:simple}, to
modal and linear type systems.
We see that modal and linear type systems apparently violate some of the nice
properties of the simply typed $\lambda$-calculus we required in
\cref{sec:simple}.
I present a solution for intuitionistic S4 modal logic, but leave a solution for
linear logic to the following chapters.

In the following two chapters, \cref{sec:semirings,sec:ren-sub-lr}, I present a
calculus $\name$ parametrised by a partially ordered semiring of \emph{usage
annotations}.
In \cref{sec:semirings}, I define the calculus, give some possible extensions,
and show that it subsumes intuitionistic S4 modal logic and Intuitionistic
Linear Logic.
In \cref{sec:ren-sub-lr}, I show that $\name$ enjoys generalised versions of the
nice properties required in \cref{sec:simple}, and I proceed to give novel
definitions of simultaneous substitutions and their action on $\name$ terms.
These two chapters are adapted from the work of \citet{WA21}.

The remaining three main chapters,
\cref{sec:framework,sec:semantics,sec:example-semantics}, adapt the syntactic
and semantic framework of \citet{AACMM21}, as presented at the end of
\cref{sec:simple}, to semiring-annotated calculi.
\Cref{sec:framework,sec:semantics} generalise the work on $\name$ presented in
\cref{sec:semirings,sec:ren-sub-lr}, respectively.
\Cref{sec:framework} shows how to formally describe the syntax of an arbitrary
semiring-annotated calculus, following the constructions used in
\cref{sec:semirings}.
\Cref{sec:semantics} then provides the generic environment-based semantic
traversal on such syntaxes, providing renaming and substitution as per
\cref{sec:ren-sub-lr} for all syntaxes as special cases of the generic
traversal.
\Cref{sec:example-semantics} then gives further example uses of the generic
traversal.

Finally, I conclude with \cref{sec:conc}, which discusses the achievements of
this thesis and openings for future work.

\section{Naming and notation conventions}

I assume familiarity with the Curry-Howard correspondence throughout this
thesis.
I make no distinction between logics and type theories, and use terminology from
each interchangeably.
Each following bullet point lists a collection of synonyms.

\begin{itemize}
  \item assumption, hypothesis, variable
  \item derivation, proof, term
  \item proposition, formula, type
  \item connective, type former
\end{itemize}

I carry out mechanised constructions and proofs in the proof assistant and
programming language Agda~\citep{Agda}.
Agda is based on Martin-L\"{o}f's intensional dependent type theory, so I
similarly present non-mechanised constructions and proofs assuming a foundation
given by dependent type theory, in a style inspired by the HoTT
Book~\citep{hottbook}.

\section{Agda primer}

I use the proof assistant and programming language Agda throughout this thesis,
with Agda code being used particularly in \cref{sec:simple} and the later
chapters.
As such, it is important for the reader to be able to read basic Agda syntax in
order to benefit from the parts of the exposition that reside in code listings.
The syntax of Agda is broadly similar to that of Haskell, and relatively close
to that of Standard ML, OCaml, and Coq's Gallina sublanguage.
I will assume that the reader is able to read basic Haskell code, and spend most
time explaining differences thereof.

>>>>

\section{Agda primer}\label{sec:agda-primer}

I use the proof assistant and programming language Agda throughout this thesis,
with Agda code being used particularly in this chapter and
\cref{sec:framework,sec:semantics,sec:example-semantics}.
As such, it is important for the reader to be able to read basic Agda syntax in
order to benefit from the parts of the exposition that reside in code listings.
The syntax of Agda is broadly similar to that of Haskell, and relatively close
to that of Standard ML, OCaml, and Coq's Gallina sublanguage.
I will assume that the reader is able to read basic Haskell code, and spend most
time explaining differences thereof.

### 2023-07-16 tex/conc.tex
[ writing ] address all remaining todos

In this thesis, I have developed a foundation for semiring-annotated calculi
presented in natural deduction style.
I have given a consolidated account of the semiring-annotated calculus $\name$,
including its relations to existing linear and modal calculi.
As part of this, I have adapted what I call \emph{bunched connectives} from
\cite{RPKV20} as a way to state the typing rules of the calculus as well as to
work with the metatheory.
The distinction between sharing and separating conjunction given by the bunched
connectives corresponds well to the notions of \emph{additive} and
\emph{multiplicative} connectives in the object language, respectively.
Following this, I have given a novel linear algebra-based definition of
\emph{environment} for semiring-annotated calculi, together with a motivation
which may serve as a basis for the corresponding definition for other
substructural systems.
The adequacy of this definition of environment is shown by my implementation of
simultaneous renaming and substitution, as well as other operations on
environments, like composition of renamings and substitutions.

With the details of $\name$ worked out, I then moved on to adapting the work of
\citet{AACMM21} so as to make it able to capture semiring-based usage
restrictions, as found in $\name$.
The syntax descriptions of the resulting system are based on the bunched
connectives, and are shown to be expressive enough to encode calculi of a
variety of forms.
In adapting the semantics, I am forced to be precise about \emph{sharing} and
\emph{separating} bunched connectives, but largely add these as a refinement of
the work of \citet{AACMM21}.
I provide the renaming and substitution operations for all expressible calculi.
I also provide more specialised examples of semantic traversals:
a usage elaborator, an NbE algorithm, a denotational semantics, and translations
between different calculi.
The usage elaborator gives an unexpected example of generic programming, which
one could not write without syntax descriptions.
% Other examples...
Together, these examples show the applicability and versatility of the framework
I have developed.

FIXME: the next two paragraphs are directly from the ESOP22 paper.

We have presented a framework for doing metatheory for a class of substructural
type systems in Agda.
The framework gives us renaming, substitution, and a usage elaborator for new
syntaxes for free, which we hope can facilitate prototyping and the
mechanisation of more interesting semantic results.
Beside the mechanised framework itself, we believe its methodology --- the use
of bunched premise combinators --- can guide and simplify the development of
(potentially unmechanised) substructural type systems.

Our account of substructurality is based on the linear algebraic
principles described by \citet{WA21}.
However, these details only really affect the definition of environment,
in which the use of linear maps is motivated by them being the standard notion
of morphism between vectors.
We could imagine that a similar notion of morphism is found for the kind of
annotations found in \citet{LicataSR17}, allowing a framework to consider
finer substructural systems.

\begin{itemize}
  \item Novel simultaneous substitution for linear type theories
  \item Generalise \citet{AACMM21} work to substructural logics
  \item Is this a reusable library? Yes, examples
  \item Methodology for metatheory of new calculi
  \item Semirings have been a reasonably robust basis.
  \item Can the code be directly generalised?
    Abstract away the algebraic structure of contexts.
\end{itemize}

@@

\begin{itemize}
  \item Fitch-style systems (can't do).
\end{itemize}

\subsection{Further work}

@@

\paragraph{Polymorphism}
An important feature of most contemporary statically typed programming languages
is polymorphism.
In particular, parametric polymorphism over types can be used to significantly
improve code reuse, and is well undestood theoretically via System $F$ and its
variants.
I have not considered polymorphism in this thesis, and neither did
\citet{AACMM21} in their paper, so whether it can be supported in the framework
presented earlier is an open question.
However, \citet{Autosubst15} did apply their related system to System $F_{<:}$,
so it may be possible to support polymorphic systems using their techniques.

@@

\subsubsection{Partiality}\label{sec:part}
\input{part.tex}

>>>>

In this thesis, I have developed a foundation for semiring-annotated calculi
presented in natural deduction style.
I have given a consolidated account of the semiring-annotated calculus $\name$,
including its relations to existing linear and modal calculi.
As part of this, I have adapted what I call \emph{bunched connectives} from
\cite{RPKV20} as a way to state the typing rules of the calculus as well as to
work with the metatheory.
The distinction between sharing and separating conjunction given by the bunched
connectives corresponds well to the notions of \emph{additive} and
\emph{multiplicative} connectives in the object language, respectively.
Following this, I have given a novel linear algebra-based definition of
\emph{environment} for semiring-annotated calculi, together with a motivation
which may serve as a basis for the corresponding definition for other
substructural systems.
The adequacy of this definition of environment is shown by my implementation of
simultaneous renaming and substitution, as well as other operations on
environments, like composition of renamings and substitutions.
The definition of environments, and hence simultaneous renaming/substitution, is
novel, but acts like the appropriate generalisation of environments for simply
typed theories.

With the details of $\name$ worked out, I then moved on to adapting the work of
\citet{AACMM21} so as to make it able to capture semiring-based usage
restrictions, as found in $\name$.
The syntax descriptions of the resulting system are based on the bunched
connectives, and are shown to be expressive enough to encode calculi of a
variety of forms.
In adapting the semantics, I am forced to be precise about \emph{sharing} and
\emph{separating} bunched connectives, but largely add these as a refinement of
the work of \citet{AACMM21}.
I provide the renaming and substitution operations for all expressible calculi.
I also provide more specialised examples of semantic traversals:
a usage elaborator, an NbE algorithm, a denotational semantics, and translations
between different calculi.
The usage elaborator gives an unexpected example of generic programming, which
one could not write without syntax descriptions.
Meanwhile, the NbE algorithm gives further justification that I have the correct
notion of environment for semiring-annotated type systems.
Together, the examples show the applicability and versatility of the framework
I have developed.

@@

To conclude, I discuss various possible future directions of the work started in
this thesis.

@@

\paragraph{Polymorphism}
An important feature of most contemporary statically typed programming languages
is polymorphism.
In particular, parametric polymorphism over types can be used to significantly
improve code reuse, and is well undestood theoretically via System $F$ and its
variants.
I have not considered polymorphism in this thesis, and neither did
\citet{AACMM21} in their paper, so whether it can be supported in the framework
presented earlier is an open question.
However, both \citet{Autosubst15} and \citet{Autosubst18} have applied similar
work to polymorphic calculi via special support, suggesting that it would be
possible to modify the framework of this thesis in a similar way.

@@

\paragraph{Partiality}\label{sec:part}
\input{part.tex}

\paragraph{A separate variable sort}
\Citet{AACMM21} mention as a limitation of their work the fact that variables
can have unrestricted type.
This limitation carries over to the work of this thesis.
In many languages --- for example, the $\mu\tilde\mu$-calculus of
\cref{sec:mu-mu-tilde} and the normal/neutral forms of a $\lambda$-calculus ---
we would like to be able to restrict the syntax so as to disallow variables of
certain kinds.
Such a restriction would make it easier to construct traversals over these
syntaxes, which I have not attempted in this thesis.

In private communication, Allais has suggested a simple solution to this problem
by which the type of variable types and the type of term types are distinct, and
related by an arbitrary relation $R$.
Each use of the \AgdaInductiveConstructor{`var} constructor then requires a
proof that $R$ relates the type of the variable picked to the type of the
conclusion desired.
For example, in the syntax of normal/neutral forms, we let the variable types be
just the types, while the term types are types tagged with whether the term is
normal or neutral.
The relation $R$ relates equal types where the term type is marked neutral.
Allais also suggests that such a scheme could be used to bind patterns in the
context, with the relatedness proof serving as a path through the pattern to a
variable of the desired type.
It remains non-trivial to work out the appropriate notion of renaming, and more
generally the appropriate notion of environment, in this setting, which I leave
to future work.

\paragraph{Logical frameworks}
I discussed in \cref{sec:lin-sys} two previous approaches to representing linear
languages in logical frameworks.
The work of this thesis should help in adapting the first approach --- making a
new logical framework with direct support for linearity --- to a broader range
of substructural calculi.
In particular, the bunched premise connectives give candidates for the required
type formers in such a logical framework.
On the other hand, the second approach --- to use an existing non-linear logical
framework but include a \texttt{linear} predicate --- should be easy to adapt to
semiring-annotated systems by replacing the \texttt{linear} predicate by a
relation between semiring elements and \texttt{term -> term} weak functions.

\paragraph{Fitch-style systems}
I mentioned in \cref{sec:fitch} an alternative way to present modal logics,
namely Fitch-style natural deduction systems.
These appear to be largely irreconcilable with the work of this thesis due to
the fact that the $\Box$-elimination rule is very sensitive to the context.
A general treatment of Fitch-style syntaxes, comparable to the treatment of
semiring-indexed syntaxes given in this thesis, would probably have to assume a
$\Box$ connective whose behaviour was determined as part of the structural rules
of the calculus.

### 2023-07-16 tex/examples.tex
[ writing ] address all remaining todos

In this chapter, I provide three example uses of semantic traversals: generic
renaming and substitution\todo{Moved}, a usage elaborator, and a denotational semantics.
The reader is also encouraged to see the far greater range of examples in the
work of \citet{AACMM21}, which should adapt to our usage-annotated setting.
Renaming and substitution are essential results, while the latter two examples
focus on usage annotations.

@@

The type of syntax descriptions depends on the type of usage annotations because
of variable binding.
For example, in the $\oc{\gr r}$-E rule of \cref{fig:lr-comb}, the right
premise binds a new variable with annotation $\gr r$, where $\gr r$ is drawn
from the ambient posemiring.
The scaling combinator also makes direct reference to the posemiring.
To produce a simplified syntax description, where usage constraints are
trivialised, we set the ambient posemiring to the 1-element $\mathbf0$
posemiring.
In contrast to syntax descriptions, even though types can contain usage
annotations, the type of types does not depend on the type of usage annotations.
This means that, in our simplified syntax, terms have types from the original
system even though variables have trivial usage annotations.
We define the $\mathbf0$ posemiring as follows, being careful to use the
0-field record type \AgdaRecord{$\top$} so that everything algebraic gets
solved by Agda's $\eta$-laws.
Indeed, in this very definition, all of the semiring operations and laws are
canonically inferred.

@@

We can now use the elaborator to automatically infer the usage
annotations for the example at the end of \cref{sec:terms}. This
allows us to write:
\ExecuteMetaData[\PaperExamplestex]{cojoin}
We have instantiated the usage elaborator so that:
\AgdaField{0\#$^{-1}$} is a singleton on $\gr0$ and $\gr\omega$, and
empty on $\gr1$; \AgdaField{1\#$^{-1}$} is a singleton on $\gr1$ and
$\gr\omega$, and empty on $\gr0$; \AgdaField{+$^{-1}$} gives $\gr0
\mapsto [(\gr0,\gr0)]$, $\gr1 \mapsto [(\gr0,\gr1),(\gr1,\gr0)]$, and
$\gr\omega \mapsto [(\gr\omega,\gr\omega)]$; and \AgdaField{*$^{-1}$}
gives $(\gr\omega, \gr0) \mapsto [\gr0]$, $(\gr\omega, \gr1) \mapsto
[]$, and $(\gr\omega, \gr\omega) \mapsto [\gr\omega]$ (omitting
$(\gr0, \_)$ and $(\gr1, \_)$ cases for brevity). Note that we do not
consider splitting $\gr\omega$ up as, say, $\gr1 + \gr\omega$, because
this splitting would introduce more non-determinism but not allow any
more terms to be typed. As such, the only non-determinism comes when
we have variables annotated $\gr1$ and need to do an additive split,
like when we apply the \AgdaInductiveConstructor{!E} rule below. At
this point, the variable can become either $\gr0$-annotated in the
left subterm and $\gr1$-annotated on the right, or vice-versa. We will
find that, because the left subterm wants to use that variable, the
former choice will be rejected. The function \AgdaFunction{var\#} is a
convenience for converting statically known natural numbers,
representing de Bruijn \emph{levels}, into variable terms.

>>>>

In this chapter, I provide four example uses of semantic traversals:
a usage elaborator, a normalisation by evaluation algorithm, a denotational
semantics, and translations between calculi.
The reader is also encouraged to see the far greater range of examples in the
work of \citet{AACMM21}, which should adapt to our usage-annotated setting.
The usage elaborator (\cref{sec:usage-elaborator}) gives an example of a program
that is generic in both syntax and usage annotations, as well as being an
essential tool for practical use of the framework.
The normalisation by evaluation algorithm (\cref{sec:nbe}) is for a specific
syntax, but exercises parts of the generic semantic traversal --- particularly,
its handling of environments --- to good effect.
The denotational semantics in terms of world-indexed relations
(\cref{sec:den-sem}) presents an interesting semantics for semiring-annotated
type systems, but with minimal use of operations provided by the framework.
Finally, the translations between $\name$ and L/nL (\cref{sec:lnl}) both justify
the encoding of L/nL given in \cref{sec:dup-lnl} and give examples of syntactic
work benefitting from the generic renaming and substitution operations.

@@

The type of syntax descriptions depends on the type of usage annotations because
of variable binding.
For example, in the $\oc{\gr r}$-E rule of \cref{fig:lr-bunched}, the right
premise binds a new variable with annotation $\gr r$, where $\gr r$ is drawn
from the ambient posemiring.
The scaling combinator also makes direct reference to the posemiring.
To produce a simplified syntax description, where usage constraints are
trivialised, we set the ambient posemiring to the 1-element $\mathbf0$
posemiring.
In contrast to syntax descriptions, even though types can contain usage
annotations, the type of types does not depend on the type of usage annotations.
This means that, in our simplified syntax, terms have types from the original
system even though variables have trivial usage annotations.
We define the $\mathbf0$ posemiring as follows, being careful to use the
0-field record type \AgdaRecord{$\top$} so that everything algebraic gets
solved by Agda's $\eta$-laws.
Indeed, in this very definition, all of the semiring operations and laws are
canonically inferred.

@@

We can now use the elaborator to automatically infer the usage
annotations for the example at the end of \cref{sec:terms-of-system}.
This allows us to write:
\ExecuteMetaData[\PaperExamplestex]{cojoin}
We have instantiated the usage elaborator so that:
\AgdaField{0\#$^{-1}$} is a singleton on $\gr0$ and $\gr\omega$, and
empty on $\gr1$; \AgdaField{1\#$^{-1}$} is a singleton on $\gr1$ and
$\gr\omega$, and empty on $\gr0$; \AgdaField{+$^{-1}$} gives $\gr0
\mapsto [(\gr0,\gr0)]$, $\gr1 \mapsto [(\gr0,\gr1),(\gr1,\gr0)]$, and
$\gr\omega \mapsto [(\gr\omega,\gr\omega)]$; and \AgdaField{*$^{-1}$}
gives $(\gr\omega, \gr0) \mapsto [\gr0]$, $(\gr\omega, \gr1) \mapsto
[]$, and $(\gr\omega, \gr\omega) \mapsto [\gr\omega]$ (omitting
$(\gr0, \_)$ and $(\gr1, \_)$ cases for brevity). Note that we do not
consider splitting $\gr\omega$ up as, say, $\gr1 + \gr\omega$, because
this splitting would introduce more non-determinism but not allow any
more terms to be typed. As such, the only non-determinism comes when
we have variables annotated $\gr1$ and need to do an additive split,
like when we apply the \AgdaInductiveConstructor{!E} rule below. At
this point, the variable can become either $\gr0$-annotated in the
left subterm and $\gr1$-annotated on the right, or vice-versa. We will
find that, because the left subterm wants to use that variable, the
former choice will be rejected. The function \AgdaFunction{var\#} is a
convenience for converting statically known natural numbers,
representing de Bruijn \emph{levels}, into variable terms.

### 2023-07-16 tex/gen-syn.tex
[ writing ] address all remaining todos

To make \AgdaFunction{sem} pass the termination check, we have three major
options:
\begin{enumerate}
  \item Assert \AgdaFunction{sem} to be terminating, bypassing the termination
    check.
  \item Use Agda's \emph{sized types} to remember that the subterms are smaller.
  \item Avoid sized types, and index terms over some user-defined type (for
    example, natural numbers or ordinal notations) which is structurally smaller
    at subterms.
\end{enumerate}

Each of these approaches has drawbacks.
Approach 1 is clearly unsafe, in the sense that the fundamental lemma
\AgdaFunction{sem} is not being completely checked for type-correctness.
Approach 2 is also unsafe, because Agda's sized type implementation is known to
make the system inconsistent~\citep{AgdaIssue1201}.
Meanwhile, approach 3 is safe, but entails a lot of manually extracted and
supplied extra arguments, which I think would distract from the presentation and
make the resulting code harder to use.
I choose to follow \citet{AACMM21} in using sized types, justified by the idea
that Agda may eventually have a sound implementation of sized types, at which
point I would want my code to be as easy to update for that new version of Agda
as possible.

>>>>

To make \AgdaFunction{sem} pass the termination check, we have four major
options:
\begin{enumerate}
  \item Assert \AgdaFunction{sem} to be terminating, bypassing the termination
    check.
  \item Use Agda's \emph{sized types} to remember that the subterms are smaller.
  \item Avoid sized types, and index terms over some user-defined type (for
    example, natural numbers or ordinal notations) which is structurally smaller
    at subterms.
  \item Inline a new, instantiated version of \AgdaFunction{map-s} wherever it
    is used.
\end{enumerate}

Each of these approaches has drawbacks.
Approach 1 is clearly unsafe, in the sense that the fundamental lemma
\AgdaFunction{sem} is not being completely checked for type-correctness.
Approach 2 is also unsafe, because Agda's sized type implementation is known to
make the system inconsistent~\citep{AgdaIssue1201}.
Meanwhile, approach 3 is safe, but entails a lot of manually extracted and
supplied extra arguments, which I think would distract from the presentation and
make the resulting code harder to use.
Finally, approach 4 is safe, but limits code reuse (both of the function
\AgdaFunction{map-s} itself and any lemmas we may prove about it).
I choose to follow \citet{AACMM21} in using sized types, justified by the idea
that Agda may eventually have a sound implementation of sized types, at which
point I would want my code to be as easy to update for that new version of Agda
as possible.
\Citet{FS22} use approach 4, and in fact have only one use of (their equivalent
of) \AgdaFunction{map-s} in their development.

### 2023-07-16 tex/ill.tex
[ writing ] address all remaining todos

In this thesis, I consider intuitionistic linear logic (ILL).
ILL can be understood as a logic of resources.
Whereas in classical logic we read a proposition $A$ to implicitly mean ``$A$ is
true'', we may read a proposition $A$ of linear logic to implicitly mean ``I
have an $A$''.
A sequent $A \vdash B$ can then be understood as saying ``if I have an $A$, I
can give it up so as to have a $B$'', and the corresponding implication
$A \multimap B$ can be understood as saying ``I have a \emph{mechanism} for
having a $B$ if I have and am willing to give up an $A$''.
\todo{Camels example}
In this section, I will carefully introduce intuitionistic linear logic so as to
make the intuitive readings more precise.

%Linear logic is a good starting point for our investigations into usage
%restrictions because it is well known, well understood, and exhibits most of the
%difficulties we will come across.
%Linear logic's lack of general weakening and contraction will be a typical
%feature of the calculi we will see, and the exponential modality $\oc$ (bang)
%will be the prototype for a range of modalities I will consider.

\subsection{Motivation of linearity}

\todo{Incorporate into chapter introduction.}
Programming applications of linearity include stateful protocols
\citep{Wadler12}, mutually exclusive capabilities (TODO: cite),
and mutation (TODO: cite).
In each of these cases, the act of using up a hypothesis allows us to divide
time into ``before'' and ``after'', with lack of duplication avoiding confusion
over which half we are in, and lack of deletion allowing us to know whether the
corresponding action actually happened.
For example, when modelling mutation, a variable refers to a single state of a
mutable value.
The \texttt{write} operation uses up such a variable, making the old state
inaccessible, and returns a new linear value to represent the new state of the
mutable value.
Such a protocol naturally also supports a \emph{freezing} operation, by which we
relinquish the ability to mutate the value in return for an immutable reference
to the value produced.

>>>>

In this thesis, I consider intuitionistic linear logic (ILL).
ILL can be understood as a logic of resources.
Whereas in classical logic we read a proposition $A$ to implicitly mean ``$A$ is
true'', we may read a proposition $A$ of linear logic to implicitly mean ``I
have an $A$''.
A sequent $A \vdash B$ can then be understood as saying ``if I have an $A$, I
can give it up so as to have a $B$'', and the corresponding implication
$A \multimap B$ can be understood as saying ``I have a \emph{mechanism} for
having a $B$ if I have and am willing to give up an $A$''.

The resource interpretation of intuitionistic linear logic is often explained by
analogy with a vending machine.
Suppose we have a vending machine which takes pound coins and sells bottles of
water for \pounds{}1 and chocolate bars for \pounds{}2.
We can represent the latter two facts by the judgements
$\textrm{\pounds}1 \vdash \mathrm{water}$ and
$\textrm{\pounds}1, \textrm{\pounds}1 \vdash \mathrm{bar}$, respectively.
Notice that having two \pounds{}1 coins is very different from having just one
\pounds{}1 coin.
We can represent having compound objects using $\otimes$-products
(tensor-products), so that having a \pounds{}1 coin together with a chocolate
bar is represented as $\textrm{\pounds}1\otimes\mathrm{bar}$, and the mechanism
for getting a chocolate bar from the vending machine is described by
$\plr{\textrm{\pounds}1 \otimes \textrm{\pounds}1} \multimap \mathrm{bar}$.
The entire protocol of one transaction with the vending machine is represented
as
$\plr{\textrm{\pounds}1 \multimap \mathrm{water}} \with
\plr{\plr{\textrm{\pounds}1 \otimes \textrm{\pounds}1} \multimap \mathrm{bar}}$,
where the $\with$-product (with-product) represents giving the user a choice of
which of the two conjuncts to have.
Finally, that we can do multiple transactions with the vending machine is
represented as
$\oc\plr{\plr{\textrm{\pounds}1 \multimap \mathrm{water}} \with
\plr{\plr{\textrm{\pounds}1 \otimes \textrm{\pounds}1} \multimap \mathrm{bar}}}
$, where the $\oc$-modality (read ``bang'' or ``of course'') represents having
an arbitrary number of copies of its argument.

Such reasoning is useful in computer science because it allows us to distinguish
what we had before some event with what we got after it.
In both intuitionistic and classical logic, if we had a \pounds{}1 coin and a
way to turn a \pounds{}1 coin into a bottle of water, we'd be able to have both
the \pounds{}1 coin and the bottle of water at once.
Instead, linear logic acknowledges that the \pounds{}1 coin was spent, so at one
time we had just the \pounds{}1 coin, and at a later time we had just the bottle
of water.
In real applications, such reasoning is useful in, for example, dealing with
mutable state~\citep{MK19} --- in which, when we mutate the state, we forget the
old value and have only the new value --- and producing a session type system~%
\citep{Wadler12} --- in which the protocol progresses as messages are sent.

In this section, I will carefully introduce intuitionistic linear logic so as to
make the intuitive readings more precise.
